---
title: "Background Information"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
always_allow_html: true
output:
  html_fragment:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results='asis', echo=FALSE, warning=FALSE, message = FALSE)
# needed for trouble shooting
boo_DEBUG <- FALSE
if(boo_DEBUG==TRUE){
  # myConfig <- file.path(system.file(package="ContDataQC"), "extdata", "config.ORIG.R")
  # source(myConfig)
}## IF ~ boo_DEBUG ~ END
```

# Background
Indices of Biotic Integrity (IBIs) were developed for macroinvertebrate and fish assemblages in wadeable and boatable, freshwater, perennial streams in Indiana. There are currently two IBIs –

* Macroinvertebrate IBI
* Fish IBI

The IBI recalibration resulted in indices for five benthic macroinvertebrate site classes and four fish site classes. The benthic macroinvertebrate site classes were defined by ecoregions and basins, including the Northern, North-Central, Southeast, South-Central, and Southwest. Note that the South-Central site class had insufficient data for IBI recalibration; therefore, we recommend users apply the Southeast IBI to sites in the South-Central site class until sufficient data are accumulated. For fish, site classes were defined by stream size, slope, and sample collection method, including the small wadeable streams, large wadeable streams, low-gradient boatable streams, and high-gradient boatable streams. 

Reports detailing the development of the macroinvertebrate and fish IBIs can be found [here](https://github.com/Blocktt/IDEMtools/tree/main/reports).

## Intended application
The IBIs are intended to be applied to samples collected from freshwater, perennial, wadeable or boatable streams in Indiana using the IDEM sampling methods.The macroinvertebrate sampling reach is defined as a 50m length of stream that includes the designated primary sampling “X-point” location (IDEM 2010). Within the reach, two sample types are collected and composited: a 10ft kicknet sample of 10 linear feet in riffle habitat (approximately 1 square meter of substrate) and a shoreline sample consisting of 25-30 sweeps along the 50m reach. Sweeps are from multiple habitats in proportion to the dominant habitat types. The fish community sampling reach is defined as 15x the average stream wetted width, with a minimum reach of 50 meters and a maximum reach of 500 meters that includes the designated “X-point”. Single pass electrofishing is performed using various standardized electrofishing methodologies and equipment depending on stream size and site accessibility. The possible list of electrofishing equipment includes backpack, longline, tote-barge or canoe, as well as a dropper anode boom with canoe for deeper pools.  The targeted habitat includes all available habitat types and should encompass at least two pool-riffle-run sequences.

If you apply the IBIs to samples that do not meet the criteria listed above, results should be interpreted with caution.

## Input metrics and scoring formulae
Macroinvertebrate IBIs were calibrated for 5 regions - the Northern (N), North-Central (NC), Southeast (SE), South-Central (SC), and Southwest (SW) (shown in Figure 1). Note that the South-Central region lacks its own IBI; therefore, apply the Southeast IBI to that region for the time being.Tables 1-4 show the input metrics and scoring formulae for the four IBIs. 

Fish IBIs were calibrated for 4 classes of streams - small wadeable streams (WS), large wadeable streams (WL), low-gradient boatable streams (BL), and high-gradient boatable streams (BH). Tables 5-8 show the input metrics and scoring formulae for the four IBIs.

### Table 1. Metrics and scoring formulae for the Northern Macroinvertebrate IBI.
```{r table_1}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab1.dir <- "Bug_N"

table1 <- read_excel(file.path(table.dir, table.file), sheet = tab1.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table1 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 2. Metrics and scoring formulae for the North-Central Macroinvertebrate IBI.
```{r table_2}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab2.dir <- "Bug_NC"

table2 <- read_excel(file.path(table.dir, table.file), sheet = tab2.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table2 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 3. Metrics and scoring formulae for the Southwest Macroinvertebrate IBI.
```{r table_3}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab3.dir <- "Bug_SW"

table3 <- read_excel(file.path(table.dir, table.file), sheet = tab3.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table3 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 4. Metrics and scoring formulae for the Southeast Macroinvertebrate IBI.The Southeast IBI is also applicable to South-Central region.
```{r table_4}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab4.dir <- "Bug_SW"

table4 <- read_excel(file.path(table.dir, table.file), sheet = tab4.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table4 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 5. Metrics and scoring formulae for the Small-wadeable fish IBI.
```{r table_5}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab5.dir <- "Fish_WS"

table5 <- read_excel(file.path(table.dir, table.file), sheet = tab5.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table5 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 6. Metrics and scoring formulae for the Large-wadeable fish IBI.
```{r table_6}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab6.dir <- "Fish_WL"

table6 <- read_excel(file.path(table.dir, table.file), sheet = tab6.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table6 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 7. Metrics and scoring formulae for the Low-gradient boatable fish IBI.
```{r table_7}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab7.dir <- "Fish_BL"

table7 <- read_excel(file.path(table.dir, table.file), sheet = tab7.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table7 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```

### Table 8. Metrics and scoring formulae for the High-gradient boatable fish IBI.
```{r table_8}
library(readxl)
library(knitr)
library(kableExtra)
# state directories
table.dir <- "tables"
table.file <- "Instruction_Tables.xlsx"
tab8.dir <- "Fish_BL"

table8 <- read_excel(file.path(table.dir, table.file), sheet = tab8.dir
                     , na = c("NA", ""), trim_ws = TRUE, skip = 0
                     , col_names = TRUE)  


table8 %>%
  kable() %>%
  kable_styling(full_width = F, position = "left")
```
